---
title: "MT5763 Assignment 1"
output: pdf_document
geometry: margin=1.5cm
header-includes:
 \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, fig.width=8, fig.height=8)
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
```

```{r loading libraries and data, include=FALSE}
library(tidyverse)
library(lubridate) #for handling dates
library(cowplot) # for plotting graphs side-by-side

dfs <- list("Seoul" = read_csv("./BikeSeoul.csv"),
            "Washington DC" = read_csv("./BikeWashingtonDC.csv"))
# since a lot of code below is repetitive, I decided to store the dataframes
# in a list for ease
```

Link to GitHub repo: <https://github.com/elizavetabektas/MT5763_1_190013200>.

```{r data wrangling, include=FALSE}

dfs[["Seoul"]] <- dfs[["Seoul"]] %>%
  
  # removing columns specified
  select(-c("Visibility (10m)",
            "Dew point temperature(C)",
            "Solar Radiation (MJ/m2)",
            "Rainfall(mm)",
            "Snowfall (cm)")
         ) %>%
  
  # filtering observations for which no bike count data was collected
  filter(`Functioning Day` == "Yes"
         )%>%
  
  # removing functioning day column as it is no longer required
  select(-c(`Functioning Day`)
         ) %>%
  
  # changing names of columns to be consistent with other dataframe
  rename(Count = `Rented Bike Count`,
         Temperature = `Temperature(C)`,
         Humidity = `Humidity(%)`,
         WindSpeed = `Wind speed (m/s)`,
         Season = Seasons
         ) %>%
  
  # converting date to a date object
  mutate(Date = dmy(Date)
         ) %>%
  
  mutate(FullDate = make_datetime(year(Date), # creating variable FullDate containing hour
                                  month(Date),
                                  mday(Date),
                                  Hour,
                                  tz="Asia/Seoul"),
         # changing factor levels of Holiday to Yes/No
         Holiday = factor(ifelse(Holiday == "Holiday", "Yes", "No"),
                          levels=c("Yes", "No")),
         # changing order of factor levels
         Season = factor(Season,
                         levels=c("Spring", "Summer", "Autumn", "Winter")))

dfs[["Washington DC"]] <- dfs[["Washington DC"]] %>%
  
  # removing specified columns
  select(-c("instant",
            "yr",
            "mnth",
            "weekday",
            "workingday",
            "weathersit",
            "atemp",
            "casual",
            "registered")
         ) %>%
  
  # changing names of columns to match the ones for Seoul
  rename(Count = cnt,
         Temperature = temp,
         Humidity = hum,
         WindSpeed = windspeed,
         Season = season,
         Hour = hr,
         Holiday = holiday,
         Date = dteday
         ) %>%
  
  mutate(Humidity = Humidity*100,              # converting Humidity to %
         Temperature = (Temperature*(39+8)-8), # converting to degrees Celcius
         WindSpeed = WindSpeed*67*1000/3600,   #  converting to m/s
         
         # changing factor levels to match Seoul
         Season = factor(recode(Season,
                         "1"="Winter",
                         "2"="Spring",
                         "3"="Summer",
                         "4"="Autumn"),
                         levels=c("Spring", "Summer", "Autumn", "Winter")),
         
         # changing factor levels of Holiday to Yes/No
         Holiday = factor(ifelse(Holiday == 1, "Yes", "No"),
                          levels=c("Yes", "No")),
         
         # converting Date to a date object
         Date = ymd(Date)) %>%
  
  # creating FullDate so it contains hour
  mutate(FullDate = make_datetime(year(Date),
                                  month(Date),
                                  mday(Date),
                                  Hour,
                                  tz="America/New_York"))
                                # Washington DC and New York have same timezone

```

# Data Visualisation
## How does air temperature varies over the course of a year?

```{r air temperature variation over a year, include=FALSE}

daily_temperature_plts <- list()

for (i in names(dfs)){
  daily_temperature_plts[[i]] <- dfs[[i]] %>%
  
  # for each day, find average air temperature
  group_by(Date) %>%
  summarize(DailyTemperature = mean(Temperature)) %>%
  
  # plot points for every day and smooth line to show trend
  ggplot(mapping = aes(x=Date, y=DailyTemperature)) +
  geom_point(color=ifelse(i=="Seoul", "orchid", "darkcyan"), size=0.5) +
  geom_smooth(color="black", method = "gam") +
  scale_x_date(date_labels = "%b %Y",       # for consistency between datasets
               breaks = ifelse(i=="Seoul", "2 months", "4 months")) +
    
  # changing default titles
  xlab("Date") +
  ylab("Temperature (°C)") +
  ggtitle(i) +
  theme(plot.title = element_text(hjust = 0.5))
}

# creating grid to display plots side-by-side
daily_temperature_plt <- plot_grid(daily_temperature_plts[["Seoul"]],
                                   daily_temperature_plts[["Washington DC"]],
                                   nrow = 2)
```

```{r output_temperature_graph, fig.cap="Daily Temperature Over Time", fig.height=5}
daily_temperature_plt
```

Seoul has data from `r min(dfs[["Seoul"]]$Date)` to `r max(dfs[["Seoul"]]$Date)`, while Washington DC data runs from `r min(dfs[["Washington DC"]]$Date)` to `r max(dfs[["Washington DC"]]$Date)`. This means the data in each dataset is from different and non-overlapping years and the total amount of data available for analysis is different.

We can see the average daily temperatures in Seoul reached just over -15°C in the winter, which was much colder than the coldest temperature of just under -5°C in Washington DC.
The warmest temperatures are similar between the cities (a few degrees over 30°C).
The coldest month in both cities is January, which can be seen both from the trendlines and individual points.
In Seoul, the warmest month is August, whilst it is July in Washington.
The cities are similar in that their daily temperature fluctuations are lower in summer compared to winter (seen from the spread of points around the trendlines).

From the gradient of the trend line, we can see the average daily temperature in Seoul increased much more rapidly between the months of February and May than between May and August. In comparison, in Washington DC, the changes in gradients were much smoother.

In the Seoul graph, we can see that the variation in average daily temperature was greater in winter compared to the summer (as the points are closer to the trendline in the summer months).
A similar trend can be seen in Washington DC, where in both years the variation in average daily air temperature is greater in winter when compared to summer, however this trend is much more subtle (the difference in point distribution around the trendline is not as obvious as in Seoul).

In Washington DC, we can see from the trendlines and individual days that January is the coldest month of every 12-month period. 

## Do seasons affect the average number of rented bikes?

```{r effect of seasons on bike demand, include=FALSE}

bar_plts <- list()
# since a lot of the plots were similar, I decided to put them all in one list

for (i in names(dfs)){
  title <- paste(i, "seasons", sep="_")
  bar_plts[[title]] <- dfs[[i]] %>%
    
  # calculating mean bikes rented in each season and error to calculate 95% CIs
  group_by(Season) %>%
  summarise(MeanRentedBikes = mean(Count),
            error = qnorm(0.975)*sd(Count)/sqrt(length(Count))) %>%
  
  # plotting bar graph with 99% CIs
  ggplot(mapping = aes(x=Season, y=MeanRentedBikes)) +
  geom_bar(stat = "identity",
           fill = ifelse(i=="Seoul", "orchid", "darkcyan")) +
  geom_errorbar(aes(ymin=MeanRentedBikes-error,
                    ymax=MeanRentedBikes+error),
                width=0.2) +
  geom_label(mapping = aes(label=round(MeanRentedBikes)), size=1.5) +
  
  # changing default titles
  ylab("Average Number of Rented Bikes per Hour") +
  ggtitle(i) +
  theme(plot.title = element_text(hjust = 0.5))
}

# as before, plotting above graphs side by side
seasons_plt<- plot_grid(bar_plts[["Seoul_seasons"]],
                         bar_plts[["Washington DC_seasons"]])

```

```{r output seasons graph, fig.cap="Effect of Seasons on Bike Renting", fig.height=7}
seasons_plt
```

It is clear that for all seasons, bike rentals are higher in Seoul compared to Washington DC.

The bike rentals have less fluctuations throughout seasons in Washington DC than in Seoul. In Seoul, the bike demand is at least 3 times higher in spring compared to winter, and at least 4 times higher in summer.

In the Seoul data, none of the 95% confidence intervals overlap, meaning the seasonal difference between bike demand is statistically significant.

This is not true for Washington DC, where the confidence intervals for spring and autumn overlap, meaning the difference is not statistically significant. The confidence intervals of summer and winter don't overlap with other seasons, meaning they are statistically significant. The bike demand in Summer is almost double that of Winter.

## Do holidays increase or decrease the demand for rented bikes?

```{r effect of holidays on bike demand}

for (i in names(dfs)){
  title <- paste(i, "holidays", sep="_")
  bar_plts[[title]] <- dfs[[i]] %>%
    
  # calculating mean bikes rented in each season and error to calculate 95% CIs
  group_by(Holiday) %>%
  summarise(MeanRentedBikes = mean(Count),
            error = qnorm(0.975)*sd(Count)/sqrt(length(Count))) %>%
  
  # plotting bar graph with 99% CIs
  ggplot(mapping = aes(x=Holiday, y=MeanRentedBikes)) +
  geom_bar(stat = "identity",
           fill = ifelse(i=="Seoul", "orchid", "darkcyan")) +
  geom_errorbar(aes(ymin=MeanRentedBikes-error,
                    ymax=MeanRentedBikes+error),
                width=0.2) +
  geom_label(mapping = aes(label=round(MeanRentedBikes)), size=1.5) +
  
  # changing default titles
  xlab("On Holiday? Yes/No") +
  ylab("Average Number of Rented Bikes") +
  ggtitle(i) +
  theme(plot.title = element_text(hjust = 0.5))
}

# as before, plotting above graphs side by side
holidays_plt_untitled <- plot_grid(bar_plts[["Seoul_holidays"]],
                         bar_plts[["Washington DC_holidays"]])
holidays_plt_title <- ggdraw() +
  draw_label("Effect of Holidays on Average Number of Rented Bikes",
             fontface = 'bold')
holidays_plt <- plot_grid(holidays_plt_title, holidays_plt_untitled, ncol = 1, rel_heights = c(0.1, 1))
```

```{r output holidays graph, fig.cap="Effect of Holidays on Average Number of Rented Bikes", fig.height=7}
holidays_plt_untitled
```

During holidays, less people are renting bikes when compared to non-holidays in both cities.

In both Seoul and Washington DC, the confidence intervals for average number of rented bikes on holidays are much wider compared to non-holidays, indicating there is greater variation in number of bikes rented during holidays. Additionally, none of the confidence intervals overlap (taking into account different y-axes), meaning these differences are statistically significant.

The relative difference between bike rentals on holidays vs non-holidays is greater in Seoul compared to Washington DC, meaning proportionally more people are renting bikes on holidays in Washington DC.

## How does the time of day affect the demand for rented bikes?

```{r effect of time of day on bike demand, include=FALSE}

for (i in names(dfs)){
  title <- paste(i, "time_of_day", sep="_")
  bar_plts[[title]] <- dfs[[i]] %>%
    
  # calculating mean bikes rented in each season and error to calculate 95% CIs
  group_by(Hour) %>%
  summarise(MeanRentedBikes = mean(Count),
            error = qnorm(0.975)*sd(Count)/sqrt(length(Count))) %>%
  
  # plotting bar graph with 99% CIs
  ggplot(mapping = aes(x=Hour, y=MeanRentedBikes)) +
  geom_bar(stat = "identity",
           fill = ifelse(i=="Seoul", "orchid", "darkcyan")) +
  geom_errorbar(aes(ymin=MeanRentedBikes-error,
                    ymax=MeanRentedBikes+error),
                width=0.2) +

  # changing default titles
  xlab("Hour") +
  ylab("Average Number of Rented Bikes") +
  ggtitle(i) +
  theme(plot.title = element_text(hjust = 0.5))
}

# as before, plotting above graphs side by side
day_time_plt <- plot_grid(bar_plts[["Seoul_time_of_day"]],
                          bar_plts[["Washington DC_time_of_day"]])
```

```{r output time of day graph, fig.cap="Effect of Holidays on Average Number of Rented Bikes", fig.height=5}
day_time_plt
```

In general, the confidence intervals in Seoul are proportionally wider than in Washington DC for every hour.

In both cities, we see a decrease in bike rental demand between 7pm and 4am. The bike demand in Washington DC drops more rapidly, whereas the relative decrease in bike usage in Seoul at night is not as great.

In Seoul, the hour with smallest average bike rentals is 4am, although this is not statistically different from 5am as the confidence intervals overlap.
Similarly, 4am has the least demand in Washington DC, and this is statistically different from every other time of day.

In Seoul, the bike demand is highest at 6pm, and this is statistically different from every other time of day.
In Washington DC, the bike demand is highest at 5pm, which is closely followed by 6pm (and the confidence intervals overlap).

In both cities, we see two clear peaks in bike demand: small peak in demand at 8am, and a high peak around 5pm-6pm. This likely corresponds to people commuting to and from work.

In Seoul, we see a steadier increase in bike demand between 10am and 6pm. In Washington DC, we see bike demand decreasing from 8am to 10am, then a small statistically insignificant peak around lunchtime 12noon and 1pm (95 % confidence intervals overlap).

## Is there an association between bike demand and the three meteorological variables (air temperature, wind speed and humidity)?

```{r effect of meteorological variables on bike demand, include=FALSE, fig.height=10}

meteorogical_plts <- list()

for (name in names(dfs)){
  temperature_title <- paste(name, "temperature", sep="_")
  temperature_plot <- dfs[[name]] %>%
    ggplot(mapping = aes(x=Temperature, y=Count)) +
    geom_point(size=0.2) +
    geom_smooth(fill='red') +
    xlab("Air Temperature (°C)") +
    ylab("Number of Bikes Rented")
  
  wind_speed_title <- paste(name, "wind_speed", sep="_")
  wind_speed_plot <- dfs[[name]] %>%
    ggplot(mapping = aes(x=WindSpeed, y=Count)) +
    geom_point(size=0.2) +
    geom_smooth(fill='red') +
    xlab("Wind Speed (m/s)") +
    ylab("Number of Bikes Rented")
  
  humidity_title <- paste(name, "humidity", sep="_")
  humidity_plot <- dfs[[name]] %>%
    ggplot(mapping = aes(x=Humidity, y=Count)) +
    geom_point(size=0.2) +
    geom_smooth(fill='red') +
    xlab("Humidity (%)") +
    ylab("Number of Bikes Rented")
  
  untitled_plot <- plot_grid(temperature_plot,
                             wind_speed_plot,
                             humidity_plot,
                             nrow = 3)
  
  plt_title <- ggdraw() + draw_label(name)
  
  meteorogical_plts[[name]] <- plot_grid(plt_title,
                                       untitled_plot,
                                       ncol=1,
                                       rel_heights = c(0.1, 1))
}

meteorology_plt_title <- ggdraw() +
  draw_label("Effect of Meteorological Variables on Bike Demand",
             fontface = 'bold')

meteorology_plt <- plot_grid(#meteorology_plt_title,
                              meteorogical_plts[["Seoul"]],
                              meteorogical_plts[["Washington DC"]],
                              ncol = 2,
                              rel_heights = c(1, 1))
```

```{r print meteorology plot, fig.cap="Effect of Meteorological Variables on Bike Demand"}
meteorology_plt
```

### Air Temperature

For both cities, we see a steady increase in bike demand as air temperature increases, then a decrease. The 95% confidence intervals are wider at the temperature extremes for Seoul and Washington DC, but we can see this is due to lack of data at the extremes.

There is likely an association between air temperature and bike demand.

### Wind Speed

In both cities, there is an initial small increase in bike demand as wind speed increases. In Seoul, the bike demand peaks at around 3m/s with a clear statistically significant peak. In Washington DC, there is no clear peak (considering the 95% confidence interval), with peak bike demand being between 4m/s and 7m/s.

We can also see that in both cities, the confidence intervals widen significantly as wind speed approaches the upper extreme.
In Washington DC, the wind speeds recorded are higher than in Seoul, indicating wind speeds in Washington DC get higher than in Seoul.

There is likely little association between bike demand and wind speed.

### Humidity

In Seoul and Washington DC, the confidence intervals are much wider at lower humidity levels (<25%).

In Seoul, bike demand remains steady as humidity increases up until 75% and begins to decrease more rapidly as humidity increases further.

In Washington DC, the peak bike demand is at around 25% humidity level, with a steady decrease in bike demand as humidity increases beyond 25%.

There is likely an association between humidity and bike demand in Washington DC, with higher humidities decreasing bike demand.

There is little association between humidity and bike demand in Seoul at lower humidities, but humidity starts having a negative effect at higher %'s.

# Statistical Modelling

To further explore the relationship between bike demand and meteorological variables, we can fit linear models.

## Fitting Linear Models

### Seoul

Below is a summary of the fitted model for the Seoul dataset.

```{r fit and summary of seoul linear model}

# fitting linear model for Seoul
seoul_lm <- lm(log(Count) ~ Season + Temperature + Humidity + WindSpeed,
               data = dfs[["Seoul"]])

summary(seoul_lm) # summary of Seoul linear model
```

From the star ratings, it is clear all coefficients (apart from the summer factor) are significant to this model. (As the p-values are all less than 0.05).

Multiple R-Squared for this model is `r summary(seoul_lm)$r.squared`, which is not close to 1, so this model does not explain the variation in the data well.

We can check our assumptions graphically:

```{r evaluating seoul linear model - histogram, fig.height=3, fig.cap="Histogram of Residuals of Seoul Model"}

# Histogram of residuals
hist(resid(seoul_lm), main = "Histogram of Residuals")
```

```{r evaluating seoul linear model - other, fig.height=7, fig.cap="Diagnostic Plots for Seoul Model"}
# Other useful plots
par(mfrow=c(2,2))
plot(seoul_lm)
par(mfrow=c(1,1))
```

Looking at __Histogram of Residuals__ and __Normal Q-Q Plot__, we can see the distribution is left-skewed. This indicates non-normality.

By looking at __Residuals vs Fitted__, we can see the red line does not follow the dashed line closely, meaning we cannot assume linearity.

In the __Residuals vs Fitted__ plot, we can see a decrease in the vertical spread of the points at higher fitted values. The red line in the __Scale-Location__ plot should be horizontal with equally randomly spread points if the assumption of constant variance holds. This indicates constant variance is questionable.

### Washington DC

Below is a summary of the fitted model for the Washington DC dataset:

```{r fit and summary of washington linear model}
washington_lm <- lm(log(Count) ~ Season + Temperature + Humidity + WindSpeed,
               data = dfs[["Washington DC"]])

summary(washington_lm)
```

All coefficients are significant in this model. In constrast to the Seoul linear model, the summer factor is significant.

Multiple R-squared is `r summary(washington_lm)$r.squared`, indicating this model does not explain well the variation seen in the data, performing worse than the Seoul model. 

```{r evaluating washington linear model - histogram, fig.height=3, fig.cap="Histogram of Residuals of Washington DC Model"}
# Histogram of residuals
hist(resid(washington_lm), main = "Histogram of Residuals")
```

```{r evaluating washington linear model - other, fig.height=7, fig.cap="Diagnostic Plots for Washington DC Model"}
# Other useful plots
par(mfrow=c(2,2))
plot(washington_lm)
par(mfrow=c(1,1))
```

From __Histogram of Residuals__ and __Normal Q-Q__ plots, we can see the distribution of residuals is even more left-skewed than that of the Seoul model. Normality does not hold.

We cannot assume linearity, as the red line in __Residuals vs Fitted__ not close to the dashed line and shows a trend at the higher fitted values.

Constant variance cannot be assumed. The spread of points around the red line in the __Residuals vs Fitted__ plot is clearly not equally random, and the red line in the __Scale-Location__ plot is clearly not horizontal.

The Washington DC linear model is clearly worse than the Seoul model due to the lower Multiple R-Squared value and clearly the assumptions don't hold.

## Calculating Confidence Intervals for Estimated Regression Coefficients

97% confidence intervals for estimated regression coefficients:

```{r CIs for estimated regression coefficients}
print("Seoul")
confint(seoul_lm, level=0.97)
print("Washington DC")
confint(washington_lm, level=0.97)
```

### Do you think these confidence intervals are reliable?

Neither model obeys the assumptions, so the confidence intervals are not reliable.

Since our assumptions don't hold, it means we cannot trust the inference that results from this model, including the confidence intervals.

## Prediction

```{r prediction}

# creating new dataframe with new data
new_data <- as_tibble(data.frame(Season = "Winter", 
                                Temperature = 0,
                                WindSpeed = 0.5,
                                Humidity = 20))

print("Seoul")
predict_seoul <- predict.lm(seoul_lm, new_data, interval="prediction", level = 0.9)
predict_seoul

print("Washington DC")
predict_washington <- predict.lm(washington_lm, new_data, interval="prediction", level = 0.9)
predict_washington
```

The point estimate from the Seoul model is `r predict_seoul[1]` with a 90% confidence interval of (`r predict_seoul[2]`, `r predict_seoul[3]`), which is not statistically different from the point estimate of `r predict_washington[1]` with CI (`r predict_washington[2]`, `r predict_washington[3]`) from the Washington DC model because their confidence intervals overlap.