---
title: "MT5763_1_190013200"
output:
  pdf_document: default
  html_document: default
---

Change size and font of plot labels for bar graphs
Change orientation of 2x3 plot?
Change LATEX formatting
Add titles to linear model plots, make the 2x2 bigger

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loading libraries and data, include=FALSE}
library(tidyverse)
library(lubridate) #for handling dates
library(cowplot) # for plotting graphs side-by-side

dfs <- list("Seoul" = read_csv("./BikeSeoul.csv"),
            "Washington DC" = read_csv("./BikeWashingtonDC.csv"))
# since a lot of code below is repetitive, I decided to store the dataframes
# in a list for ease
```

## Data Wrangling
`BikeSeoul.csv` and `BikeWashingtonDC.csv` were read in and cleaned for downstream analysis.
__Note__: if running the code in this report, the __.csv__ files need to be in the same directory as the __.Rmd__ file.

```{r data wrangling, include=FALSE}

dfs[["Seoul"]] <- dfs[["Seoul"]] %>%
  
  # removing columns specified
  select(-c("Visibility (10m)",
            "Dew point temperature(C)",
            "Solar Radiation (MJ/m2)",
            "Rainfall(mm)",
            "Snowfall (cm)")
         ) %>%
  
  # filtering observations for which no bike count data was collected
  filter(`Functioning Day` == "Yes"
         )%>%
  
  # removing functioning day column as it is no longer required
  select(-c(`Functioning Day`)
         ) %>%
  
  # changing names of columns to be consistent with other dataframe
  rename(Count = `Rented Bike Count`,
         Temperature = `Temperature(C)`,
         Humidity = `Humidity(%)`,
         WindSpeed = `Wind speed (m/s)`,
         Season = Seasons
         ) %>%
  
  # converting date to a date object
  mutate(Date = dmy(Date)
         ) %>%
  
  mutate(FullDate = make_datetime(year(Date), # creating variable FullDate containing hour
                                  month(Date),
                                  mday(Date),
                                  Hour,
                                  tz="Asia/Seoul"),
         # changing factor levels of Holiday to Yes/No
         Holiday = factor(ifelse(Holiday == "Holiday", "Yes", "No"),
                          levels=c("Yes", "No")),
         # changing order of factor levels
         Season = factor(Season,
                         levels=c("Spring", "Summer", "Autumn", "Winter")))

dfs[["Washington DC"]] <- dfs[["Washington DC"]] %>%
  
  # removing specified columns
  select(-c("instant",
            "yr",
            "mnth",
            "weekday",
            "workingday",
            "weathersit",
            "atemp",
            "casual",
            "registered")
         ) %>%
  
  # changing names of columns to match the ones for Seoul
  rename(Count = cnt,
         Temperature = temp,
         Humidity = hum,
         WindSpeed = windspeed,
         Season = season,
         Hour = hr,
         Holiday = holiday,
         Date = dteday
         ) %>%
  
  mutate(Humidity = Humidity*100,              # converting Humidity to %
         Temperature = (Temperature*(39+8)-8), # converting to degrees Celcius
         WindSpeed = WindSpeed*67*1000/3600,   #  converting to m/s
         
         # changing factor levels to match Seoul
         Season = factor(recode(Season,
                         "1"="Winter",
                         "2"="Spring",
                         "3"="Summer",
                         "4"="Autumn"),
                         levels=c("Spring", "Summer", "Autumn", "Winter")),
         
         # changing factor levels of Holiday to Yes/No
         Holiday = factor(ifelse(Holiday == 1, "Yes", "No"),
                          levels=c("Yes", "No")),
         
         # converting Date to a date object
         Date = ymd(Date)) %>%
  
  # creating FullDate so it contains hour
  mutate(FullDate = make_datetime(year(Date),
                                  month(Date),
                                  mday(Date),
                                  Hour,
                                  tz="America/New_York"))
                                # Washington DC and New York have same timezone

```

# Data Visualisation
## How does air temperature varies over the course of a year?
We have the air temperature values for every hour of every day.
This means there will be fluctuations in temperature in each 24-hour period.
This is why I found average daily temperature prior to creating plots.

```{r air temperature variation over a year, include=FALSE}

daily_temperature_plts <- list()

for (i in names(dfs)){
  daily_temperature_plts[[i]] <- dfs[[i]] %>%
  
  # for each day, find average air temperature
  group_by(Date) %>%
  summarize(DailyTemperature = mean(Temperature)) %>%
  
  # plot points for every day and smooth line to show trend
  ggplot(mapping = aes(x=Date, y=DailyTemperature)) +
  geom_point(color=ifelse(i=="Seoul", "orchid", "darkcyan")) +
  geom_smooth(color="black", method = "gam") +
  scale_x_date(date_labels = "%b %Y",       # for consistency between datasets
               breaks = ifelse(i=="Seoul", "2 months", "4 months")) +       
    
  # changing default titles
  xlab("Date") +
  ylab("Temperature (C)") +
  ggtitle(i) +
  theme(plot.title = element_text(hjust = 0.5))
}

# I decided to create two separate graphs, then show them side-by-side using
# cowplot's plot_grid because:
# 1. Seoul has 1 year of data, Washington has 2 years of data
#    To comment on year-on-year changes, I didn't want to average the two years.
# 2. Seoul data is from 2011, Washington is 2017 and 2018.
#    So I couldn't use the same x-axis without only retaining month data.

# creating grid to display plots side-by-side
daily_temperature_plt_untitled <- plot_grid(daily_temperature_plts[["Seoul"]],
                                   daily_temperature_plts[["Washington DC"]],
                                   nrow = 2)
daily_temperature_plt_title <- ggdraw() +
  draw_label("Fluctuations in Daily Temperature",
             fontface = 'bold')

daily_temperature_plt <- plot_grid(daily_temperature_plt_title,
                        daily_temperature_plt_untitled,
                        ncol = 1,
                        rel_heights = c(0.1, 1))
```
```{r output graph, fig.width=8, fig.height=8}
daily_temperature_plt
```

## Washington DC
In general, the daily temperature fluclations between the months of May and August are smaller than in all other months across the two years.
From the trendlines and individual days, January is the coldest month of every 12-month period. From the data available, we can see December 2010 is much colder than December 2011, but this needs further investigation as December 2010 data is incomplete. There is less variation in summer 2011 than summer 2012, with summer 2012 being on average a few degrees warmer on average.

## Comparing the Cities
Seoul has data from `r min(dfs[["Seoul"]]$Date)` to `r max(dfs[["Seoul"]]$Date)`, while Washington DC data runs from `r min(dfs[["Washington DC"]]$Date)` to `r max(dfs[["Washington DC"]]$Date)`. This means the data in each dataset is from different and non-overlapping years and the amount of data is different.

We can see the average daily temperatures in Seoul reached just over -15°C in the winter, which was much colder than the coldest temperature of just under -5°C in Washington DC.

The warmest temperatures are similar between the cities (a few degrees over 30°C).

The coldest month in both cities is January, which can be seen both from the trendlines and individual points.
In Seoul, the warmest month is August, whilst it is July in Washington.

The cities are similar in that their daily temperature fluctuations are lower in summer compared to winter (seen from the spread of points around the trendlines).

From the gradient of the trend line, we can see the average daily temperature in Seoul increased much more rapidly between the months of February and May than between May and August. In comparison, in Washington DC, the changes in gradient were much smoother.

## Do seasons affect the average number of rented bikes?

The plots below show the average number of rented bikes per hour for each season, with error bars representing 95% condifence intervals.

```{r effect of seasons on bike demand}

bar_plts <- list()
# since a lot of the plots were similar, I decided to put them all in one list

for (i in names(dfs)){
  title <- paste(i, "seasons", sep="_")
  bar_plts[[title]] <- dfs[[i]] %>%
    
  # calculating mean bikes rented in each season and error to calculate 99% CIs
  group_by(Season) %>%
  summarise(MeanRentedBikes = mean(Count),
            error = qnorm(0.975)*sd(Count)/sqrt(length(Count))) %>%
  
  # plotting bar graph with 99% CIs
  ggplot(mapping = aes(x=Season, y=MeanRentedBikes)) +
  geom_bar(stat = "identity",
           fill = ifelse(i=="Seoul", "orchid", "darkcyan")) +
  geom_errorbar(aes(ymin=MeanRentedBikes-error,
                    ymax=MeanRentedBikes+error),
                width=0.2) +
  geom_text(mapping = aes(label=round(MeanRentedBikes)), size=3) +
  
  # changing default titles
  ylab("Average Number of Rented Bikes per Hour") +
  ggtitle(i) +
  theme(plot.title = element_text(hjust = 0.5))
}

# as before, plotting above graphs side by side
seasons_plt <- plot_grid(bar_plts[["Seoul_seasons"]],
                         bar_plts[["Washington DC_seasons"]])
seasons_plt_title <- ggdraw() +
  draw_label("Effect of Seasons on Bike Renting",
             fontface = 'bold')
plot_grid(seasons_plt_title, seasons_plt, ncol = 1, rel_heights = c(0.1, 1))

```

From these graphs, we can see that for all seasons, bike rentals are higher in Seoul compared to Washington DC.

The bike rentals have less fluctuations throughout seasons in Washington DC than in Seoul. In Seoul, the bike demand is at least 3 times higher in spring compared to winter, and at least 4 times higher in summer.

In the Seoul data, none of the error bars overlap, meaning the seasonal difference between bike demand is statistically significant.
This is not true for Washington DC, where the confidence intervals for spring and autumn overlap, meaning the difference is not statistically significant. The confidence intervals of summer and winter don't overlap with other seasons, meaning they are statistically significant. The bike demand in Summer is almost double that of Winter.

## Do holidays increase or decrease the demand for rented bikes?

The plots below show the average number of rented bikes per hour against whether they were rented on a holiday, with error bars representing 95% condifence intervals.

```{r effect of holidays on bike demand}
######## QUESTION: Do holidays increase or decrease the demand for rented bikes?

for (i in names(dfs)){
  title <- paste(i, "holidays", sep="_")
  bar_plts[[title]] <- dfs[[i]] %>%
    
  # calculating mean bikes rented in each season and error to calculate 99% CIs
  group_by(Holiday) %>%
  summarise(MeanRentedBikes = mean(Count),
            error = qnorm(0.975)*sd(Count)/sqrt(length(Count))) %>%
  
  # plotting bar graph with 99% CIs
  ggplot(mapping = aes(x=Holiday, y=MeanRentedBikes)) +
  geom_bar(stat = "identity",
           fill = ifelse(i=="Seoul", "orchid", "darkcyan")) +
  geom_errorbar(aes(ymin=MeanRentedBikes-error,
                    ymax=MeanRentedBikes+error),
                width=0.2) +
  geom_text(mapping = aes(label=round(MeanRentedBikes)), size=3) +
  
  # changing default titles
  xlab("On Holiday? Yes/No") +
  ylab("Average Number of Rented Bikes") +
  ggtitle(i) +
  theme(plot.title = element_text(hjust = 0.5))
}

# as before, plotting above graphs side by side
holidays_plt <- plot_grid(bar_plts[["Seoul_holidays"]],
                         bar_plts[["Washington DC_holidays"]])
holidays_plt_title <- ggdraw() +
  draw_label("Effect of Holidays on Average Number of Rented Bikes",
             fontface = 'bold')
plot_grid(holidays_plt_title, holidays_plt, ncol = 1, rel_heights = c(0.1, 1))
```

During holidays, less people are renting bikes when compared to not during holidays.

The confidence intervals for number of bike rentals on holidays are much wider compared to non-holidays, indicating there is greater variation in number of bikes rented during holidays. Additionally, the confidence intervals don't overlap, meaning the differences are statistically significant.

The relative difference between bike rentals on holiday vs not on holiday is greater in Seoul compared to Washington DC, meaning proportionally more people are renting bikes on holidays in Washington DC.

## How does the time of day affect the demand for rented bikes?

The plots below show the average number of rented bikes per hour against time of day, with error bars representing 95% condifence intervals.

```{r effect of time of day on bike demand}

for (i in names(dfs)){
  title <- paste(i, "time_of_day", sep="_")
  bar_plts[[title]] <- dfs[[i]] %>%
    
  # calculating mean bikes rented in each season and error to calculate 99% CIs
  group_by(Hour) %>%
  summarise(MeanRentedBikes = mean(Count),
            error = qnorm(0.975)*sd(Count)/sqrt(length(Count))) %>%
  
  # plotting bar graph with 99% CIs
  ggplot(mapping = aes(x=Hour, y=MeanRentedBikes)) +
  geom_bar(stat = "identity",
           fill = ifelse(i=="Seoul", "orchid", "darkcyan")) +
  geom_errorbar(aes(ymin=MeanRentedBikes-error,
                    ymax=MeanRentedBikes+error),
                width=0.2) +

  # changing default titles
  xlab("Hour") +
  ylab("Average Number of Rented Bikes") +
  ggtitle(i) +
  theme(plot.title = element_text(hjust = 0.5))
}

# as before, plotting above graphs side by side
day_time_plt <- plot_grid(bar_plts[["Seoul_time_of_day"]],
                          bar_plts[["Washington DC_time_of_day"]])
day_time_plt_title <- ggdraw() +
  draw_label("Effect of Holidays on Average Number of Rented Bikes",
             fontface = 'bold')
plot_grid(day_time_plt_title, day_time_plt, ncol = 1, rel_heights = c(0.1, 1))
```

In general, the confidence intervals in Seoul are wider than in Washington DC for every hour.

In both cities, we see a decrease in bike rental demand between 7pm and 4am. The bike demand in Washington DC drops more rapidly, whereas the relative decrease in bike usage in Seoul at night is not as big.

In Seoul, the hour with smallest average bike rentals is 4am, although this is not statistically different from 5am as the confidence intervals overlap.
Similarly, 4am has the least demand in Washington DC, but this is statistically different from every other time of day.

In Seoul, the bike demand is highest at 6pm, and this is statistically different from every other time of day.
In Washington DC, the bike demand is highest at 5pm, which is closely followed by 6pm (and the confidence intervals overlap).

In both cities, we see a small peak in demand at 8am, after which it decreases.

In Seoul, we see a steadier increase in bike demand between 10 am and 6pm. In Washington DC, we see bike demand decreasing from 8am to 10am, then a small statistically insignificant peak around lunchtime 12noon and 1pm (error bars overlapping).

## Is there an association between bike demand and the three meteorological variables (air temperature, wind speed and humidity)?

The plots below show the effect of three meteorological variables on bike demand.

```{r effect of meteorological variables on bike demand, include=FALSE}

meteorogical_plts <- list()

for (name in names(dfs)){
  temperature_title <- paste(name, "temperature", sep="_")
  temperature_plot <- dfs[[name]] %>%
    ggplot() +
    geom_smooth(mapping = aes(x=Temperature, y=Count)) +
    xlab("Air Temperature (C)") +
    ylab("Number of Bikes Rented")
  
  wind_speed_title <- paste(name, "wind_speed", sep="_")
  wind_speed_plot <- dfs[[name]] %>%
    ggplot() +
    geom_smooth(mapping = aes(x=WindSpeed, y=Count)) +
    xlab("Wind Speed (m/s)") +
    ylab("Number of Bikes Rented")
  
  humidity_title <- paste(name, "humidity", sep="_")
  humidity_plot <- dfs[[name]] %>%
    ggplot() +
    geom_smooth(mapping = aes(x=Humidity, y=Count)) +
    xlab("Humidity (%)") +
    ylab("Number of Bikes Rented")
  
  untitled_plot <- plot_grid(temperature_plot,
                             wind_speed_plot,
                             humidity_plot,
                             ncol = 3)
  
  plt_title <- ggdraw() + draw_label(name)
  
  meteorogical_plts[[name]] <- plot_grid(plt_title,
                                       untitled_plot,
                                       ncol=1,
                                       rel_heights = c(0.1, 1))
}

meteorology_plt_title <- ggdraw() +
  draw_label("Effect of Meteorological Variables on Bike Demand",
             fontface = 'bold')

meteorology_plt <- plot_grid(meteorology_plt_title,
                              meteorogical_plts[["Seoul"]],
                              meteorogical_plts[["Washington DC"]],
                              ncol = 1,
                              rel_heights = c(0.1, 1, 1))
```

```{r print meteorology plot}
meteorology_plt
```

### Air Temperature

For both cities, we see a steady increase in bike demand as air temperature increases, then decrease, following a similar trend. The 95% confidence intervals widen at the temperature extremes.

### Wind Speed

In both cities, there is an initial increase in bike demand as wind speed increases. In Seoul, the bike demand peaks at around 3m/s with a clear statistically significant peak. In Washington DC, there is no clear peak (considering the 95% confidence interval), with peak bike demand being between 4m/s and 7m/s.

We can also see that in both cities, the confidence intervals widen significantly as wind speed approaches the upper extreme.
The wind speeds in Washington DC get higher than in Seoul.

### Humidity

In Seoul, the confidence interval is much wider at lower humidity levels (<25%) than other humidity levels in Seoul.
The confidence interval in Seoul is wider than in Washington DC.

In both cities, we see that as humidity increases, bike demand initially increases. As humidity levels keep increading, bike demand decreases. Despite following the same general trend, the shapes of the trend lines look quite different.
In Washington DC, the peak bike demand is at around 25% humidity level, with lowest demand being at 0% humidity level, which is not statistically different from 100% humidity level.
On the other hand, in Seoul the peak happens at just under 50% humidity, and the bike demand at extremes being statistically different. Bike demand is lowest at 100% humidity.

# Statistical Modelling

## Fitting Linear Models

A linear model was fitted with log count as outcome, and season, air temperature, humidity and wind speed as predictors.

### Seoul

Below is a summary of the fitted model for the Seoul dataset.

```{r fit and summary of seoul linear model}

# fitting linear model for Seoul
seoul_lm <- lm(log(Count) ~ Season + Temperature + Humidity + WindSpeed,
               data = dfs[["Seoul"]])

summary(seoul_lm) # summary of Seoul linear model
```

From the star ratings, it is clear all coefficients (apart from the summer season) are significant to this model. (As the p-values are all less than 0.05)

Multiple R-Squared for this model is 0.4941, which is not close to 1, so this model does not explain the variation in the data well.

The linear regression model is

$Y_i$ ~ $N(\alpha + \beta*x_i)$ for $i=1, ..., n$ with $Y_1, ..., Y_n$ independent

Thus when fitting the normal linear regression model, we are making the assumptions:

1. $Y_1, . . . , Y_n$ are independent,
2. $Y_1, . . . , Y_n$ are normally distributed,
3. The mean of $Y_i$'s a linear function of xi,
4. $Y_1, . . . , Y_n$ have the same variance.

We can check these assumptions graphically:

```{r evaluating seoul linear model - histogram}

# Histogram of residuals
hist(resid(seoul_lm))
```

```{r evaluating seoul linear model - other}
# Other useful plots
par(mfrow=c(2,2))
plot(seoul_lm)
par(mfrow=c(1,1))
```

Looking at __Normal Q-Q plot__ and __Histogram of resid(seoul_lm)__, we can see the distribution is left-skewed. This indicates non-normality.

To investigate linearity, we can look at __Residuals vs Fitted__. The red line is close to the dashed line, indicating linearity holds reasonably well.

In the __Residuals vs Fitted__ plot, we can see a decrease in the vertical spread of the points at higher fitted values. The red line in the __Scale-Location__ plot should be horizontal with equally randomly spread points if the assumption of constant variance holds. This indicates constant variance is questionable.

### Washington DC

Below is a summary of the fitted model for the Washington DC dataset.

```{r fit and summary of washington linear model}
washington_lm <- lm(log(Count) ~ Season + Temperature + Humidity + WindSpeed,
               data = dfs[["Washington DC"]])

summary(washington_lm)
```
All coefficients are significant in this model. In constrast to the Seoul linear model, summer is significant.

Multiple R-squared is 0.278, indicating this model does not explain well the variation seen in the data, performing worse than the Seoul model. 

Checking the assumptions for this linear model graphically:

```{r evaluating washington linear model - histogram}
# Histogram of residuals
hist(resid(washington_lm))
```

```{r evaluating washington linear model - other}
# Other useful plots
par(mfrow=c(2,2))
plot(washington_lm)
par(mfrow=c(1,1))
```

From __Histogram of resid(washington_lm)__ and __Normal Q-Q__ plots, we can see the distribution of residuals is even more left-skewed than that of the Seoul model. Normality does not hold.

We can assume linearity, as the red line in __Residuals vs Fitted__ is quite close to the dashed line. 

Constant variance cannot be assumed, however. The spread of points around the red line in the __Residuals vs Fitted__ plot is clearly not equally random, and the red line in the __Scale-Location__ plot is clearly not horizontal.

The Washington DC linear model is clearly worse than the Seoul model due to the lower Multiple R-Squared value and clearly less assumptions hold.

## Calculating Confidence Intervals for Estimated Regression Coefficients

97% confidence intervals for estimated regression coefficients:

```{r CIs for estimated regression coefficients}
print("Seoul")
confint(seoul_lm, level=0.97)
print("Washington DC")
confint(washington_lm, level=0.97)
```

### Do you think these confidence intervals are reliable?

We know that SeasonSummer is the only not significant coefficient in the Seoul model, therefore its confidence interval is too wide to be useful.

Neither model obeys the assumptions, so the confidence intervals are not reliable.

## Prediction

```{r prediction}

# creating new dataframe with new data
new_data <- as_tibble(data.frame(Season = "Winter", 
                                Temperature = 0,
                                WindSpeed = 0.5,
                                Humidity = 20))

print("Seoul")
predict_seoul <- predict.lm(seoul_lm, new_data, interval="prediction", level = 0.9)
predict_seoul

print("Washington DC")
predict_washington <- predict.lm(washington_lm, new_data, interval="prediction", level = 0.9)
predict_washington
```

The point estimate from the Seoul model is `r predict_seoul[1]` with a 90% confidence interval of (`r predict_seoul[2]`, `r predict_seoul[3]`), which is not statistically different from the point estimate of `r predict_washington[1]` with CI (`r predict_washington[2]`, `r predict_washington[3]`) from the Washington DC model because their confidence intervals overlap.

